Vertical and oblique cultural transmission fluctuating in time and in space

### 2018-07-05 ms/ms_sunnyvale.tex
small corrections; fixed phrasing at end of appendix

% Title page
\title{Vertical and oblique {\color{red}cultural} transmission fluctuating in time and in space}

@@

In evolutionary genetics, the properties of transmission are known to be important, e.g., uniparental vs.\ biparental; haploid vs.\ diploid; with or without recombination; sexual or asexual. For cultural evolution, the mode of transmission between individuals of the same or different generations is also central to the dynamics of cultural traits.
%Important modes of cultural transmission are vertical (from parent to offspring), horizontal (between peers), and oblique (from adults in the parental generation to offspring) transmission. 
{\color{red}Using analysis with epidemic theory}, \citet[ch.~3]{Cavalli-Sforza1981}  introduced models in which a specific cultural trait is transmitted either vertically, that is, directly from parent to offspring, obliquely, that is, from non-parental members of the parent's generation to an offspring{\color{red}, or horizontally, that is, among members of the same generational cohort.}
Mathematical models of cultural evolution may include one or more modes of cultural transmission and in most of these models the mode of transmission by individuals is fixed; that is, it does not vary over time.
For example, \cite{Feldman1996}, \cite{Wakano2004}, \cite{Aoki2005}, and \citet{McElreath2008} focused on competition between individual learning, innate learning, and social learning; \citet{Fogarty2017} compared the effects of different oblique mechanisms (random, success-biased, best-of-$k$, one-to-many) on the cultural richness and diversity of the population; and \citet{Aoki2012} modeled scenarios in which individual and social learning occur during separate stages in life.
In these studies, the cultural transmission rule did not change during the evolutionary process.

@@

\begin{figure*}[h]
\centering
\includegraphics[width=0.5\linewidth]{../figures/{transmission}.png}
\caption{
\textbf{Cultural transmission with mixed vertical and oblique transmission.}
When a newborn matures, she copies the phenotype---color---of her mother with probability $\rho$, therefore becoming blue, or from some other female with probability $1-\rho$, in which case her color will depend on the frequency of blue and red adult females.}
\label{fig:transmission}
\end{figure*}

% TODO what kind of spatial differences can we imagine in transmission?
% two populations, one has first cousin marriages and one doesn't (endogmia)
% orthodox religious communities
% boarding schools for one population
% foraging in groups or families
% pigmies women 'migration' - adoption by husband family
% interesting: migration correlated with transmission rate, future work

% matrilineal primates - vertical transmission stronger in females than males
% similar differences in ultra religious human societies

@@

\paragraph{Symmetric periods---$AkBk$.}
We first consider environments that fluctuate periodically every $k$ generations between favoring $A$ and $B$.
Using a linear approximation, fixations of $A$ ($x^*=1$) and $B$ ($x^*=0$) are locally stable if, respectively,
\begin{equation}\begin{aligned}
\big[F'_A(1) F'_B(1)\big]^k < 1,  &\quad \text{and}\\
\big[F'_A(0) F'_B(0)\big]^k < 1. \\
\end{aligned}\end{equation}
By definition, a \emph{protected polymorphism} exists if neither fixation is stable~\citep{Prout1968}, that is, if neither of these conditions are met.
Therefore, for a protected polymorphism to exist, we require
\begin{equation}
1 < F'_A(1) F'_B(1) = \Big(1-\rho_A\frac{s}{1+s}\Big)\Big(1+\rho_B s\Big) 
= 1+\frac{s}{1+s}\big[\rho_B(1+s-s\rho_A)-\rho_A\big],
\end{equation}
and
\begin{equation}
1 < F'_A(0) F'_B(0) = \Big(1+\rho_A s\Big)\Big(1-\rho_B \frac{s}{1+s}\Big) 
= 1+\frac{s}{1+s}\big[\rho_A(1+s-s\rho_B)-\rho_B\big],
\end{equation}
which can be summarized as a condition on $\rho_B$,
\begin{equation} \label{eq:poly_condition_periodic_fluc}
\frac{\rho_A}{1+s(1-\rho_A)} < \rho_B < \frac{(1+s)\rho_A}{1+s\rho_A},
\end{equation}
or as a condition on the difference between $\rho_B$ and $\rho_A$
\begin{equation}
-s\rho_B(1-\rho_A) < \rho_B - \rho_A < s\rho_A(1-\rho_B).
\end{equation}
We can state this as Result 1.

@@

\paragraph{Period of one---$A1B1$.}
In the case  $k=1$ we can find $x^*$ the stable frequency of $A$ at the protected polymorphism.
Set $F_B(F_A(x))-x = G(x) \cdot x (1-x) \cdot \gamma$ with
\begin{equation} \begin{aligned} \label{eq:xstar_periodic_fluc_k=1}
&G(x) = ax^2+bx+c, \\
&\gamma = -\frac{1 + sx}{s} \Big[ 1 + s + s^2 x (1-\rho)(1+x)\Big], \\
&a = s^2 \rho_A (1-\rho_A) (1 - \rho_B + \rho_B/\rho_A)) > 0, \\
&b = s(1-\rho_A)(2\rho_B - s\rho_A(1-\rho_B)), \\
& c = \rho_B - \rho_A - s\rho_A(1-\rho_B).
\end{aligned} \end{equation}
Then $x^*$ is a solution of the quadratic $G(x)=0$.
The condition \eqref{eq:poly_condition_periodic_fluc}, which guarantees that $0 < x^* < 1$, is equivalent to $c<0$, and therefore it also guarantees that $\sqrt{b^2-4ac} > b$.
So, if $0<s<1$ (so that $b$ is guaranteed to be positive if LHS of eq.~\ref{eq:poly_condition_periodic_fluc} holds), then 
\begin{equation}
x^*= \frac{-b+\sqrt{b^2-4ac}}{2a}.
\end{equation}
\autoref{fig:rho1_rho2_phase_k=1} shows $x^*$ and highlights the area of the parameter space in which a protected polymorphism exists.
The figure demonstrates that the stronger the selection (x-axis), the greater the fluctuations in $\rho$  (y-axis) that still allow a polymorphic population (area between the dashed lines).

%\begin{align}
%G(0) = c > 0  \Leftrightarrow \rho_B > \frac{\rho_A(1+s)}{1+\rho_As}, \\
%G'(0) = b > 0 \Leftrightarrow \rho_B > \frac{\rho_A s}{2 + \rho_As}, \\
%G(1) = a + b + c = (1+s)[\rho_B - \rho_A + s \rho_B (1-\rho_A)], \\
%G'(1) = 2a + b = s(1-\rho_A)(2(1+s)\rho_B + s\rho_A(1-\rho_B)) > 0.
%\end{align}

\begin{figure*}[htb]
\centering
\includegraphics[width=0.65\linewidth]{../figures/{rho1_rho2_phase_k=1}.pdf}
\caption{
\textbf{Protected polymorphism.}
The stable equilibrium of the frequency of phenotype $A$ (eq.~\ref{eq:xstar_periodic_fluc_k=1}) for different selection coefficients ($s$ on x-axis) and size of fluctuations in vertical transmission rates ($\rho_B-\rho_A$ on y-axis) when both selection and transmission fluctuate every generation ($k=1$).
Dashed lines represent $\rho_B=\frac{\rho_A}{1+s(1-\rho_A)}$ and $\rho_B=\frac{(1+s)\rho_A}{1+s\rho_A}$, the limits on $\rho_B-\rho_A$ from {\color{red}inequalities~(\ref{eq:poly_condition_periodic_fluc})} that permit a protected polymorphism.
Here, $\rho_A=0.5$.}
\label{fig:rho1_rho2_phase_k=1}
\end{figure*}

\paragraph{Asymmetric period---$AkBl$.}
More generally, phenotype $A$ could be favored for $k$ generations and phenotype $B$ for $l$ generations, with the transmission rate following the same cycle with $\rho=\rho_A$ when $A$ is favored and $\rho=\rho_B$ when $B$ is favored.
The requirements for a protected polymorphism are now
\begin{equation}\begin{aligned}
& 1 < (F'_A(0))^k (F'_B(0))^l = a^k \big(b - s \Delta \rho /(1+s)\big)^l, \quad \text{and} \\
& 1 < (F'_A(1))^k (F'_B(1))^l = b^k (a + s \Delta \rho)^l,
\end{aligned}\end{equation}
where $a=1+\rho_A s>1$, $b=1-\rho_A\frac{s}{1+s}<1$, and $\Delta \rho = \rho_B - \rho_A$.
This leads to a condition similar to eq.~20 in~\citet{Ram2018}, but more complex due to the inclusion of $\Delta \rho$:
\begin{equation} \label{eq:poly_condition_periodic_fluc_k_l}
\frac{-\log{b}}{\log{\big(a + s \Delta \rho \big)}} < 
\frac{l}{k} < 
\frac{\log{a}}{-\log{\big(b - s \Delta \rho/(1+s)\big)}}.
\end{equation}
Therefore, for a given value of $\rho_A$, if $\rho_B>\rho_A$ then increasing the vertical transmission rate of $B$ decreases the environmental period ratio $l/k$ that permits a protected polymorphism; decreasing $\rho_B$  will have the opposite effect: it increases the ratio $l/k$ that permits a protected polymorphism.

@@

Second, if $s_t$ and $\rho_t$ are not independent such that $cov(\rho_t, s_t) \ne 0$, a fixation can occur.
Let $w_{A,t}, w_{B,t} \sim U(0,1)$, $s_t=(w_{A,t}-w_{B,t})/w_{B,t}$ and $\rho_t \sim Beta(1+s_t, 1)$ (a beta distribution with parameters $1+s_t$ and $1$). 
The covariance of $s_t$ and $\rho_t$ is positive ($cov(s_t, \rho_t) \approx 4$ {\color{red}as} estimated by averaging over $10^8$ random values of $s_t$ and $\rho_t$); that is, vertical transmission is more likely when $A$ is favored (i.e. $s_t>0$) and oblique transmission is more  likely when $B$ is favored (i.e. $s_t<0$; \autoref{fig:beta}).
Then $E[log(1+z_t)] >0$ and $B$ almost surely doesn't fix. 
Also, $E[log(1-z_t / (1+s_t)] < 0$, so fixation of $A$ is stochastically locally stable.
The opposite occurs if $\rho_t \sim \beta(1, 1+s_t)$ and the covariance of $s_t$ and $\rho_t$ is negative ($cov(s_t, \rho_t) \approx -4$). In that case, fixation of $B$ is stochastically locally stable and $A$ almost surely doesn't fix.

\begin{figure*}[hbt]
\centering
\includegraphics[width=\linewidth]{../figures/{beta}.png}
\caption{
\textbf{Covariance of selection and transmission.}
\textbf{(A)} Histogram of $w_{A,t}/w_{B,t}$ where $w_{A,t}$ and $w_{B,t}$ are identically and independently distributed uniform random variables $U(0,1)$.
\textbf{(B)} Histogram of $s_t = (w_{A,t}-w_{B,t})/w_{B,t}$.
\textbf{(C)} Histogram of $\rho_t \sim Beta(1+s_t, 1)$.
\textbf{(D)} The joint distribution of $\rho_t$ and $s_t$ demonstrates a positive correlation $cov(s_t, \rho_t)>0$.
}
\label{fig:beta}
\end{figure*}

@@

\begin{figure*}[hbt]
\centering
\includegraphics[width=0.75\linewidth]{../figures/{rho1_rho2_stoch_p}.pdf}
\caption{
\textbf{Stochastic local stability.}
Here, $s_t=0.05$ and $\rho_t=\rho_1$ with probability $p=0.505$ and $s_t=-0.05$ and $\rho_t=\rho_2$ with probability $1-p=0.495$.
The diagonal represents the case of no transmission fluctuations; \citet[Fig.~2]{Ram2018} demonstrated that with a constant transmission rate $\rho=0.1$ and the above distribution of $s_t$, neither fixation is stochastically stable.
}
\label{fig:rho1_rho2_stoch_p}
\end{figure*}

@@

We obtain a result similar to result 11 from~\citet{Ram2018}:
The mean $\mu(x)$ and variance $\sigma^2(x)$ of the change in the frequency $x$ of phenotype $A$ in the case of a cycling environment $AkBl$, where $k+l=n$, are
\begin{equation} \label{eq:drift_diffusion_terms}
\mu(x) = S_n x(1-x),
\quad \text{and} \quad
\sigma^2(x) = n x (1-x),
\end{equation}
where $S_n = \sum_{t=1}^{n}{z_t}$ and $z_t = \rho_t s_t$.
Furthermore, combining eq.~\ref{eq:drift_diffusion_terms} with eqs.~46-47 from \citet{Ram2018}, we find that the probability {\color{red}of} fixation of phenotype $A$ when its initial frequency if $x$, is
\begin{equation}
u(x) = \frac{1 - e^{-2 \frac{S_n}{n} x}}{1 - e^{-2 \frac{S_n}{n}}}.
\end{equation}
From result 10 in \citet{Ram2018}, $u(x)$ is monotone increasing in $S_n/n$ which is the average selection coefficient of $A$ weighted by the vertical transmission rates $\rho_t$.
Therefore, if $s_t$ and $\rho_t$ are positively (negatively) correlated, $S_n/n$ increases (decreases), and the fixation probability $u(x)$ increases (decreases). 
This occurs because selection  affects only those individuals that transmit their phenotype to their own offspring (i.e. vertically), and a fraction $1-\rho_t$ of the population is effectively masked (for better or worse) from selection at each generation.

@@

Following migration~\citep{Karlin1981}, the frequencies of $A$ in the two demes are
\begin{equation} \label{eq:migration_model_migration}
\begin{aligned}
x_1' &= (1-m_1) x_1^s + m_1 x_2^s, \\
x_2' &= m_2 x_1^s + (1-m_2) x_2^s,
\end{aligned}
\end{equation}
where $0 \le m_1, m_2 \le 1/2$ are the migration rates, such that $m_1$ is the fraction of the population of deme 1 replaced by individuals from deme 2, and vice versa for $m_2$.
This is a \emph{two-deme stepping-stone migration scheme}~\citep[][eq.~2.17]{Karlin1982}.
Analysis of this general model is difficult, though some analytical results may be attained~\citep[see~Principle~6.1]{Karlin1982}.

\paragraph{Asymmetric migration and symmetric selection.}

\autoref{fig:asym_migration} shows some numerical results for the case of symmetric selection symmetric selection $w_1=1/w_2=w>1$.
We focus on composite parameters of the model:
$m_1/m_2$, on the x-axis, is the ratio of migration rates into deme 1 and deme 2; when this ratio is large, deme 1 accepts more migrants than deme 2.
$\rho_1/\rho_2$, on the y-axis the ratio of the vertical transmission rates in deme 1 and deme 2; when this ratio is large, individuals in deme 1 use vertical transmission more often then individuals in deme 2.

The results demonstrate that fixation of phenotype $A$ is stable if migration to and oblique transmission within deme 1, where $A$ is favored, are higher than in deme 2.
The {\color{red}opposite} is true for phenotype $B$.
A protected polymorphism exists if neither fixation is stable: if migration ratios are positively correlated -- vertical transmission occurs more often in the deme that accepts more migrants -- or if both ratios are close to unity, that is, the differences between the demes in terms of migration and transmission are small (\autoref{fig:asym_migration}).

\begin{figure*}[hbt]
\centering
\includegraphics[width=0.75\linewidth]{../figures/asym_migration.pdf}
\caption{
\textbf{Oblique transmission and asymmetric migration.}
Classification of the stable equilibrium of the system in eqs.~\ref{eq:migration_model_selection_transmission}-\ref{eq:migration_model_migration} for different ratios of the migration rates (x-axis) and vertical transmission rates (y-axis) in the two demes.
Stability was determined for 10,000 random choices of $m_1$, $m_2$, $\rho_1$, and $\rho_2$ by calculating the leading eigenvalue of the Jacobian of the system.
Blue markers denote cases in which the leading eigenvalue of the Jacobian at $x_1=x_2=0$ was less than 1, leading to fixation of $B$.
Red markers denote cases in which the leading eigenvalue of the Jacobian at $x_1=x_2=1$ was less than 1, leading to fixation of $A$.
Green markers denote cases in which both leading eigenvalues were larger than 1, leading to a protected polymorphism.
Here, the fitness values are $w_1=1/w_2=2$.
}
\label{fig:asym_migration}
\end{figure*}

\paragraph{Symmetric migration and selection.}

@@

We have the following results:
\begin{itemize}
\item With only oblique transmission ($\rho_1=\rho_2=0$), {\color{red}there are only} neutral equilibria $(x^*,x^*)$ for any $0 \le x^* \le 1$.
\item With only vertical transmission ($\rho_1=\rho_2=1$), the fixation equilibria  $(0,0)$, $(1,1)$ are unstable and there exists a protected polymorphism
\begin{equation}
\begin{aligned} \label{eq:migration_model_unconditional_symmetric_polymorphism}
x_1^* &= \frac{w-1-m(w+1) + \sqrt{\Delta}}{2(w-1)}, \\
x_2^* &= 1-x_1^*,
\end{aligned}
\end{equation}
where $\Delta = m^2(w+1)^2+(1-2m)(w-1)^2$.
\item With only vertical transmission in one deme ($\rho_1=1$) and a combination of both vertical and oblique transmission in the other deme ($\rho_2=\rho$), fixation of $B$ is unstable, and fixation of $A$ is stable if and only if the vertical transmission rate in deme 2 is below $\hat \rho$; that is,
\begin{equation} \label{eq:migration_model_unconditional_symmetric_condition_rho}
\rho < \hat \rho = \frac{m}{m+(1-m)(w-1)} < 1,
\end{equation}
or if the migration rate is above $\hat m$; that is,
\begin{equation} \label{eq:migration_model_unconditional_symmetric_condition_m}
m > \hat m = \frac{\rho w - 1}{\rho w + 1}.
\end{equation}
\end{itemize}
{\color{red}The proofs of (22) and (23) are in Appendix I.}
\paragraph{Examples.}

@@

Comparing our results with previous results from a similar model with a fixed rate of vertical transmission~\citep{Ram2018}, we find that a protected polymorphism can be maintained only if fluctuations in the rate of vertical transmission are small, and that stronger selection on the transmitted trait permits larger fluctuations in the rate of transmission while still maintaining a protected polymorphism. 
In the case of fluctuations in space, the greater the separation between the two populations {\color{red}(i.e., the smaller the migration rates)} the larger {\color{red} can be the fluctuations that maintain} polymorphism; however, as migration becomes more frequent, even small differences in the vertical transmission rate will eliminate the polymorphism.
When fluctuations are stochastic, we find that if vertical transmission covaries with selection, the phenotype that has a higher probability of being vertically transmitted  when it is favored will likely eventually fix in the population.
However, if transmission and selection are independent, then a polymorphism can be maintained if selection does not, on average, favor one phenotype over the other.

>>>>

% Title page
\title{Vertical and oblique cultural transmission fluctuating in time and in space}

@@

In evolutionary genetics, the properties of transmission are known to be important, e.g., uniparental vs.\ biparental; haploid vs.\ diploid; with or without recombination; sexual or asexual. For cultural evolution, the mode of transmission between individuals of the same or different generations is also central to the dynamics of cultural traits.
Using analysis with epidemic theory, \citet[ch.~3]{Cavalli-Sforza1981}  introduced models in which a specific cultural trait is transmitted either vertically, that is, directly from parent to offspring, obliquely, that is, from non-parental members of the parent's generation to an offspring, or horizontally, that is, among members of the same generational cohort.
Mathematical models of cultural evolution may include one or more modes of cultural transmission and in most of these models the mode of transmission by individuals is fixed; that is, it does not vary over time {\color{red}or space}.
For example, \cite{Feldman1996}, \cite{Wakano2004}, \cite{Aoki2005}, and \citet{McElreath2008} focused on competition between individual learning, innate learning, and social learning; \citet{Fogarty2017} compared the effects of different oblique mechanisms (random, success-biased, best-of-$k$, one-to-many) on the cultural richness and diversity of the population; and \citet{Aoki2012} modeled scenarios in which individual and social learning occur during separate stages in life.
In these studies, the cultural transmission rule did not change during the evolutionary process.

@@

% figure
\begin{figure*}[h]
\centering
\includegraphics[width=0.5\linewidth]{../figures/{transmission}.png}
\caption{
\textbf{Cultural transmission with mixed vertical and oblique transmission.}
When a newborn matures, she copies the phenotype---color---of her mother with probability $\rho$, therefore becoming blue, or from some other female with probability $1-\rho$, in which case her color will depend on the frequency of blue and red adult females.}
\label{fig:transmission}
\end{figure*}

@@

% Symmetric periods
\paragraph{Symmetric periods---$AkBk$.}
We first consider environments that fluctuate periodically every $k$ generations between favoring $A$ and $B$.
Using a linear approximation, fixations of $A$ ($x^*=1$) and $B$ ($x^*=0$) are locally stable if, respectively,
\begin{equation}\begin{aligned}
\big[F'_A(1) F'_B(1)\big]^k < 1,  &\quad \text{and}\\
\big[F'_A(0) F'_B(0)\big]^k < 1. \\
\end{aligned}\end{equation}
By definition, a \emph{protected polymorphism} exists if neither fixation is stable~\citep{Prout1968}, that is, if neither of these conditions are met.
Therefore, for a protected polymorphism to exist, we require
\begin{equation}
1 < F'_A(1) F'_B(1) = \Big(1-\rho_A\frac{s}{1+s}\Big)\Big(1+\rho_B s\Big) 
= 1+\frac{s}{1+s}\big[\rho_B(1+s-s\rho_A)-\rho_A\big],
\end{equation}
and
\begin{equation}
1 < F'_A(0) F'_B(0) = \Big(1+\rho_A s\Big)\Big(1-\rho_B \frac{s}{1+s}\Big) 
= 1+\frac{s}{1+s}\big[\rho_A(1+s-s\rho_B)-\rho_B\big],
\end{equation}
which can be summarized as a condition on $\rho_B$,
\begin{equation} \label{eq:poly_condition_periodic_fluc}
\frac{\rho_A}{1+s(1-\rho_A)} < \rho_B < \frac{(1+s)\rho_A}{1+s\rho_A},
\end{equation}
or as a condition on the difference between $\rho_B$ and $\rho_A$
\begin{equation}
-s\rho_B(1-\rho_A) < \rho_B - \rho_A < s\rho_A(1-\rho_B).
\end{equation}
We can state this as Result 1.

@@

% Period of one
\paragraph{Period of one---$A1B1$.}
In the case  $k=1$ we can find $x^*$ the stable frequency of $A$ at the protected polymorphism.
Set $F_B(F_A(x))-x = G(x) \cdot x (1-x) \cdot \gamma$ with
\begin{equation} \begin{aligned} \label{eq:xstar_periodic_fluc_k=1}
&G(x) = ax^2+bx+c, \\
&\gamma = -\frac{1 + sx}{s} \Big[ 1 + s + s^2 x (1-\rho)(1+x)\Big], \\
&a = s^2 \rho_A (1-\rho_A) (1 - \rho_B + \rho_B/\rho_A)) > 0, \\
&b = s(1-\rho_A)(2\rho_B - s\rho_A(1-\rho_B)), \\
& c = \rho_B - \rho_A - s\rho_A(1-\rho_B).
\end{aligned} \end{equation}
Then $x^*$ is a solution of the quadratic $G(x)=0$.
The condition \eqref{eq:poly_condition_periodic_fluc}, which guarantees that $0 < x^* < 1$, is equivalent to $c<0$, and therefore it also guarantees that $\sqrt{b^2-4ac} > b$.
So, if $0<s<1$ (so that $b$ is guaranteed to be positive if LHS of eq.~\ref{eq:poly_condition_periodic_fluc} holds), then 
\begin{equation}
x^*= \frac{-b+\sqrt{b^2-4ac}}{2a}.
\end{equation}
\autoref{fig:rho1_rho2_phase_k=1} shows $x^*$ and highlights the area of the parameter space in which a protected polymorphism exists.
The figure demonstrates that the stronger the selection (x-axis), the greater the fluctuations in $\rho$  (y-axis) that still allow a polymorphic population (area between the dashed lines).

% figure
\begin{figure*}[htb]
\centering
\includegraphics[width=0.65\linewidth]{../figures/{rho1_rho2_phase_k=1}.pdf}
\caption{
\textbf{Protected polymorphism.}
The stable equilibrium of the frequency of phenotype $A$ (eq.~\ref{eq:xstar_periodic_fluc_k=1}) for different selection coefficients ($s$ on x-axis) and size of fluctuations in vertical transmission rates ($\rho_B-\rho_A$ on y-axis) when both selection and transmission fluctuate every generation ($k=1$).
Dashed lines represent $\rho_B=\frac{\rho_A}{1+s(1-\rho_A)}$ and $\rho_B=\frac{(1+s)\rho_A}{1+s\rho_A}$, the limits on $\rho_B-\rho_A$ from inequalities~\eqref{eq:poly_condition_periodic_fluc} that permit a protected polymorphism.
Here, $\rho_A=0.5$.}
\label{fig:rho1_rho2_phase_k=1}
\end{figure*}

% Asymmetric period
\paragraph{Asymmetric period---$AkBl$.}
More generally, phenotype $A$ could be favored for $k$ generations and phenotype $B$ for $l$ generations, with the transmission rate following the same cycle with $\rho=\rho_A$ when $A$ is favored and $\rho=\rho_B$ when $B$ is favored.
The requirements for a protected polymorphism are now
\begin{equation}\begin{aligned}
& 1 < (F'_A(0))^k (F'_B(0))^l = a^k \big(b - s \Delta \rho /(1+s)\big)^l, \quad \text{and} \\
& 1 < (F'_A(1))^k (F'_B(1))^l = b^k (a + s \Delta \rho)^l,
\end{aligned}\end{equation}
where $a=1+\rho_A s>1$, $b=1-\rho_A\frac{s}{1+s}<1$, and $\Delta \rho = \rho_B - \rho_A$.
This leads to a condition similar to eq.~20 in~\citet{Ram2018}, but more complex due to the inclusion of $\Delta \rho$:
\begin{equation} \label{eq:poly_condition_periodic_fluc_k_l}
\frac{-\log{b}}{\log{\big(a + s \Delta \rho \big)}} < 
\frac{l}{k} < 
\frac{\log{a}}{-\log{\big(b - s \Delta \rho/(1+s)\big)}}.
\end{equation}
Therefore, for a given value of $\rho_A$, if $\rho_B>\rho_A$ then increasing the vertical transmission rate of $B$ decreases the environmental period ratio $l/k$ that permits a protected polymorphism; decreasing $\rho_B$  will have the opposite effect: it increases the ratio $l/k$ that permits a protected polymorphism.

@@

Second, if $s_t$ and $\rho_t$ are not independent such that $cov(\rho_t, s_t) \ne 0$, a fixation can occur.
Let $w_{A,t}, w_{B,t} \sim U(0,1)$, $s_t=(w_{A,t}-w_{B,t})/w_{B,t}$ and $\rho_t \sim Beta(1+s_t, 1)$ (a beta distribution with parameters $1+s_t$ and $1$). 
The covariance of $s_t$ and $\rho_t$ is positive ($cov(s_t, \rho_t) \approx 4$ as estimated by averaging over $10^8$ random values of $s_t$ and $\rho_t$); that is, vertical transmission is more likely when $A$ is favored (i.e. $s_t>0$) and oblique transmission is more  likely when $B$ is favored (i.e. $s_t<0$; \autoref{fig:beta}).
Then $E[log(1+z_t)] >0$ and $B$ almost surely doesn't fix. 
Also, $E[log(1-z_t / (1+s_t)] < 0$, so fixation of $A$ is stochastically locally stable.
The opposite occurs if $\rho_t \sim \beta(1, 1+s_t)$ and the covariance of $s_t$ and $\rho_t$ is negative ($cov(s_t, \rho_t) \approx -4$). In that case, fixation of $B$ is stochastically locally stable and $A$ almost surely doesn't fix.

% figure
\begin{figure*}[hbt]
\centering
\includegraphics[width=\linewidth]{../figures/{beta}.png}
\caption{
\textbf{Covariance of selection and transmission.}
\textbf{(A)} Histogram of $w_{A,t}/w_{B,t}$ where $w_{A,t}$ and $w_{B,t}$ are identically and independently distributed uniform random variables $U(0,1)$.
\textbf{(B)} Histogram of $s_t = (w_{A,t}-w_{B,t})/w_{B,t}$.
\textbf{(C)} Histogram of $\rho_t \sim Beta(1+s_t, 1)$.
\textbf{(D)} The joint distribution of $\rho_t$ and $s_t$ demonstrates a positive correlation $cov(s_t, \rho_t)>0$.
}
\label{fig:beta}
\end{figure*}

@@

% figure
\begin{figure*}[hbt]
\centering
\includegraphics[width=0.75\linewidth]{../figures/{rho1_rho2_stoch_p}.pdf}
\caption{
\textbf{Stochastic local stability.}
Here, $s_t=0.05$ and $\rho_t=\rho_1$ with probability $p=0.505$ and $s_t=-0.05$ and $\rho_t=\rho_2$ with probability $1-p=0.495$.
The diagonal represents the case of no transmission fluctuations; \citet[Fig.~2]{Ram2018} demonstrated that with a constant transmission rate $\rho=0.1$ and the above distribution of $s_t$, neither fixation is stochastically stable.
}
\label{fig:rho1_rho2_stoch_p}
\end{figure*}

@@

We obtain a result similar to result 11 from~\citet{Ram2018}:
The mean $\mu(x)$ and variance $\sigma^2(x)$ of the change in the frequency $x$ of phenotype $A$ in the case of a cycling environment $AkBl$, where $k+l=n$, are
\begin{equation} \label{eq:drift_diffusion_terms}
\mu(x) = S_n x(1-x),
\quad \text{and} \quad
\sigma^2(x) = n x (1-x),
\end{equation}
where $S_n = \sum_{t=1}^{n}{z_t}$ and $z_t = \rho_t s_t$.
Furthermore, combining eq.~\ref{eq:drift_diffusion_terms} with eqs.~46-47 from \citet{Ram2018}, we find that the probability of fixation of phenotype $A$ when its initial frequency if $x$, is
\begin{equation}
u(x) = \frac{1 - e^{-2 \frac{S_n}{n} x}}{1 - e^{-2 \frac{S_n}{n}}}.
\end{equation}
From result 10 in \citet{Ram2018}, $u(x)$ is monotone increasing in $S_n/n$ which is the average selection coefficient of $A$ weighted by the vertical transmission rates $\rho_t$.
Therefore, if $s_t$ and $\rho_t$ are positively (negatively) correlated, $S_n/n$ increases (decreases), and the fixation probability $u(x)$ increases (decreases). 
This occurs because selection  affects only those individuals that transmit their phenotype to their own offspring (i.e. vertically), and a fraction $1-\rho_t$ of the population is effectively masked (for better or worse) from selection at each generation.

@@

Following migration, the frequencies of $A$ in the two demes are
\begin{equation} \label{eq:migration_model_migration}
\begin{aligned}
x_1' &= (1-m_1) x_1^s + m_1 x_2^s, \\
x_2' &= m_2 x_1^s + (1-m_2) x_2^s,
\end{aligned}
\end{equation}
where $0 \le m_1, m_2 \le 1/2$ are the migration rates, such that $m_1$ is the fraction of the population of deme 1 replaced by individuals from deme 2, and vice versa for $m_2$.
This is a \emph{two-deme stepping-stone migration scheme}~\citep[][eq.~2.17]{Karlin1982}.
Analysis of this general model is difficult, though some analytical results may be attained~\citep[see~Principle~6.1]{Karlin1982}.

% Asymmetric migration and symmetric selection
\paragraph{Asymmetric migration and symmetric selection.}

\autoref{fig:asym_migration} shows some numerical results for the case of symmetric selection{\color{red},} $w_1=1/w_2=w>1$.
We focus on composite parameters of the model:
$m_1/m_2$, on the x-axis, is the ratio of migration rates into deme 1 and deme 2; when this ratio is large, deme 1 accepts more migrants than deme 2.
$\rho_1/\rho_2$, on the y-axis{\color{red},} the ratio of the vertical transmission rates in deme 1 and deme 2; when this ratio is large, individuals in deme 1 use vertical transmission more often then individuals in deme 2.

The results demonstrate that fixation of phenotype $A$ is stable if migration to and oblique transmission within deme 1, where $A$ is favored, are higher than in deme 2.
The opposite is true for phenotype $B$.
A protected polymorphism exists if neither fixation is stable: if migration ratios are positively correlated -- vertical transmission occurs more often in the deme that accepts more migrants -- or if both ratios are close to unity, that is, the differences between the demes in terms of migration and transmission are small (\autoref{fig:asym_migration}).

% figure
\begin{figure*}[hbt]
\centering
\includegraphics[width=0.75\linewidth]{../figures/asym_migration.pdf}
\caption{
\textbf{Oblique transmission and asymmetric migration.}
Classification of the stable equilibrium of the system in eqs.~\ref{eq:migration_model_selection_transmission}-\ref{eq:migration_model_migration} for different ratios of the migration rates (x-axis) and vertical transmission rates (y-axis) in the two demes.
Stability was determined for 10,000 random choices of $m_1$, $m_2$, $\rho_1$, and $\rho_2$ by calculating the leading eigenvalue of the Jacobian of the system.
Blue markers denote cases in which the leading eigenvalue of the Jacobian at $x_1=x_2=0$ was less than 1, leading to fixation of $B$.
Red markers denote cases in which the leading eigenvalue of the Jacobian at $x_1=x_2=1$ was less than 1, leading to fixation of $A$.
Green markers denote cases in which both leading eigenvalues were larger than 1, leading to a protected polymorphism.
Here, the fitness values are $w_1=1/w_2=2$.
}
\label{fig:asym_migration}
\end{figure*}

% Symmetric migration and selection
\paragraph{Symmetric migration and selection.}

@@

We have the following results:
\begin{itemize}
\item With only oblique transmission ($\rho_1=\rho_2=0$), there are only neutral equilibria $(x^*,x^*)$ for any $0 \le x^* \le 1$.
\item With only vertical transmission ($\rho_1=\rho_2=1$), the fixation equilibria  $(0,0)$, $(1,1)$ are unstable and there exists a protected polymorphism
\begin{equation}
\begin{aligned} \label{eq:migration_model_unconditional_symmetric_polymorphism}
x_1^* &= \frac{w-1-m(w+1) + \sqrt{\Delta}}{2(w-1)}, \\
x_2^* &= 1-x_1^*,
\end{aligned}
\end{equation}
where $\Delta = m^2(w+1)^2+(1-2m)(w-1)^2$.
\item With only vertical transmission in one deme ($\rho_1=1$) and a combination of both vertical and oblique transmission in the other deme ($\rho_2=\rho$), fixation of $B$ is unstable, and fixation of $A$ is stable if and only if the vertical transmission rate in deme 2 is below $\hat \rho$; that is,
\begin{equation} \label{eq:migration_model_unconditional_symmetric_condition_rho}
\rho < \hat \rho = \frac{m}{m+(1-m)(w-1)} < 1,
\end{equation}
or if the migration rate is above $\hat m$; that is,
\begin{equation} \label{eq:migration_model_unconditional_symmetric_condition_m}
m > \hat m = \frac{\rho w - 1}{\rho w + 1}.
\end{equation}
\end{itemize}
{\color{red}The proofs of (22) and (23) are in Appendix I.}
\paragraph{Examples.}

@@

Comparing our results with previous results from a similar model with a fixed rate of vertical transmission~\citep{Ram2018}, we find that a protected polymorphism can be maintained only if fluctuations in the rate of vertical transmission are small, and that stronger selection on the transmitted trait permits larger fluctuations in the rate of transmission while still maintaining a protected polymorphism. 
In the case of fluctuations in space, the greater the separation between the two populations (i.e., the smaller the migration rates) the larger can be the fluctuations that maintain polymorphism; however, as migration becomes more frequent, even small differences in the vertical transmission rate will eliminate the polymorphism.
When fluctuations are stochastic, we find that if vertical transmission covaries with selection, the phenotype that has a higher probability of being vertically transmitted  when it is favored will likely eventually fix in the population.
However, if transmission and selection are independent, then a polymorphism can be maintained if selection does not, on average, favor one phenotype over the other.
